\textbf{The Illusion Game: A Novel Experimental Paradigm Provides Evidence in Favour of a General Factor of Visual Illusion Sensitivity and Personality Correlates}

### 2022-09-07 manuscript.Rmd
address comments by steph

One key challenge hindering the further development of illusion research is the relative difficulty in adapting visual illusions to an experimental setting, which typically requires the controlled modulation of the specific variables of interest. To address this issue, we first developed a parametric framework to manipulate visual illusions, which we implemented and made accessible in the open-source software *Pyllusion* [@makowski2021]. This software allows us to generate different types of classic visual illusions (e.g., Müller-Lyer, Ponzo, Delboeuf, Ebbinghaus, ...) with a continuous and independent modulation of two parameters: *illusion strength* and *task difficulty* (see **Figure 1**).

@@

The final sample included 46 participants (Mean age = 26.7, SD = 7.7, range: [19, 60]; Sex: 39.1% females, 56.5% males).

@@

We then extracted the inter-individual variability in the effect of illusion strength and its interaction with task difficulty, and used it as participant-level scores. Finally, We explored the relationship of these indices across different illusions using exploratory factor analysis (EFA) and structural equation modelling (SEM).

@@

```{r message=FALSE, warning=FALSE, out.width="100%", fig.cap="Top: the effect of illusion strength and task difficulty on the error rage and reaction time (RT) for each individual illusions. The solid line represent the General Additive Model (GAM), and the dashed line correspond to its approximation via linear models. Descriptive data is shown with stacked dots (errors are hanging from the top) and distributions for RTs. Negative values for illusion strength correspond to congruent (i.e., facilitating) illusion effects. Task difficulty (the objective difference between the targets of perceptual decision) levels are shown as colors, with lowest values corresponding to harder trials. Each illusion type is surrounded by 4 extreme examples of stimuli, corresponding to the hardest difficutly (on top) and the strongest illusion (on the right for incongruent illusions). Bottom: We extracted the effect slope of the illusion strength and its interaction with task difficulty for each participant. We fitted a Structural Equation Model (SEM) suggesting that these manifest variables group to first-level illusion-specific latent factors, which then load on a general factor of illusion sensitivity (Factor \\textit{i})."}
knitr::include_graphics("figures/Figure2.png")
```

@@

Though imperfect, we believe that the random-slope models capture inter-individual differences with more accuracy (and are also more conservative estimates due shrinkage) than basic empirical scores, such as the total number of errors, or the average RT. Thus, for each illusion and within each participant, we extracted the effect of illusion strength and its interaction with task difficulty when the illusion effect was incongruent. These twenty participant-level scores were subjected to exploratory factor analysis (EFA). The Method Agreement Procedure [@parametersArticle] suggested the presence of 7 latent factors. An oblique (*oblimin* rotation) factor solution explaining 66.69% of variance suggested separate dimensions for the effect of Zöllner, White, Poggendorff, Contrast, Ebbinghaus, Delboeuf, and a common factor for the parameters related to Müller-Lyer, Vertical-Horizontal, Ponzo and Rod and Frame. We submitted these factors to a second-level analysis and extracted two orthogonal (*varimax* rotation) factors. The first factor was loaded by all the previous dimensions with the exception of Delboeuf, which formed its own separate factor.

@@

The effect on errors was monotonic for most illusions, with the exception of Delboeuf and Zöllner. For both of them, mildly congruent illusion strengths (which theoretically were supposed to be associated with less errors than incongruent effects) were related to small and strong increases of errors, respectively. For the Delboeuf illusion, we believe that this was due to an artifact caused by the illusion generation algorithm: the outline of the target circles was always created as slightly bigger, which made the difference between them more obvious at an illusion strength of 0. This was fixed in latest release of *Pyllusion* (v1.2), which now generate outlines of the same size as the target circle. For the Zöllner illusion, the observed non-monotonic pattern is actually consistent with previous reports [@kitaoka2000; @kitaoka2007], suggesting an acute angle contraction effect at very small as well as at sufficiently large angles (below 10 degrees for the former and between 50 to 90 degrees for the latter) between the target horizontal line and the biasing horizontal bars when the illusion strength is weak.

@@

```{r message=FALSE, warning=FALSE, fig.cap="The upper plots show the distribution of demographic and dispositional variables. The middle plots show the relationship between illusion sensitivty scores, sex, and age (solid lines indicate significant relationships). Bottom plots show the correlation between the general factor of illusion sensitivity (Factor \\textit{i}), and personality traits.", out.width="100%"}
knitr::include_graphics("figures/Figure3.png")
```

@@

We report significant links between inter-individual indices of illusion sensitivity and varialbes related to experimental context, demographic characteristics and personality. Firstly, screen size was found to have a significant negative relationship with the sensitivity to the two contrast-based illusions, namely the White and Contrast illusions. One possible explanation can be found in the mechanism by which visual systems filter through more low spatial frequencies when the size of the target object is small [@dixon2014]. As this filtering process excludes illumination information from visual processing, smaller screen sizes could yield artifactual changes in brightness perception, which in turn could attenuate the illusory effect of luminance-related illusions.

>>>>

One key challenge hindering the further development of illusion research is the relative difficulty of adapting visual illusions to an experimental setting, which typically requires the controlled modulation of the specific variables of interest. To address this issue, we first developed a parametric framework to manipulate visual illusions, which we implemented and made accessible in the open-source software *Pyllusion* [@makowski2021]. This software allows us to generate different types of classic visual illusions with a continuous and independent modulation of two parameters: *illusion strength* and *task difficulty* (see **Figure 1**).

@@

The final sample included 46 participants (Mean age = 26.7, SD = 7.7, range: [19, 60]; Sex: 39.1% females, 56.5% males, and 4.4% other).

@@

We then extracted the inter-individual variability in the effect of illusion strength and its interaction with task difficulty, and used it as participant-level scores. Finally, we explored the relationship of these indices across different illusions using exploratory factor analysis (EFA) and structural equation modelling (SEM).

@@

```{r message=FALSE, warning=FALSE, out.width="100%", fig.cap="Top: the effect of illusion strength and task difficulty on the error rate and reaction time (RT) for each individual illusions. The solid line represent the General Additive Model (GAM), and the dashed line correspond to its approximation via linear models. Descriptive data is shown with stacked dots (errors are hanging from the top) and distributions for RTs. Negative values for illusion strength correspond to congruent (i.e., facilitating) illusion effects. Task difficulty (the objective difference between the targets of perceptual decision) levels are shown as colors, with lowest values corresponding to harder trials. Each illusion type is surrounded by 4 extreme examples of stimuli, corresponding to the hardest difficulty (on top) and the strongest illusion (on the right for incongruent illusions). Bottom: We extracted the effect slope of the illusion strength and its interaction with task difficulty for each participant. We fitted a Structural Equation Model (SEM) suggesting that these manifest variables group to first-level illusion-specific latent factors, which then load on a general factor of illusion sensitivity (Factor \\textit{i})."}
knitr::include_graphics("figures/Figure2.png")
```

@@

Though imperfect, we believe that the random-slope models capture inter-individual differences with more accuracy (and are also more conservative estimates due to shrinkage) than basic empirical scores, such as the total number of errors, or the average RT. Thus, for each illusion and within each participant, we extracted the effect of illusion strength and its interaction with task difficulty when the illusion effect was incongruent. These twenty participant-level scores were subjected to exploratory factor analysis (EFA). The Method Agreement Procedure [@parametersArticle] suggested the presence of 7 latent factors. An oblique (*oblimin* rotation) factor solution explaining 66.69% of variance suggested separate dimensions for the effect of Zöllner, White, Poggendorff, Contrast, Ebbinghaus, Delboeuf, and a common factor for the parameters related to Müller-Lyer, Vertical-Horizontal, Ponzo and Rod and Frame. We submitted these factors to a second-level analysis and extracted two orthogonal (*varimax* rotation) factors. The first factor was loaded by all the previous dimensions with the exception of Delboeuf, which formed its own separate factor.

@@

The effect on errors was monotonic for most illusions, with the exception of Delboeuf and Zöllner. For both of them, mildly congruent illusion strengths (which theoretically were supposed to be associated with less errors than incongruent effects) were related to small and strong increases of errors, respectively. For the Delboeuf illusion, we believe that this was due to an artifact caused by the illusion generation algorithm: the outline of the target circles was always created as slightly bigger, which made the difference between them more obvious at an illusion strength of 0. This was fixed in latest release of *Pyllusion* (v1.2), which now generate outlines of the same size as the target circle. For the Zöllner illusion, the observed non-monotonic pattern is actually consistent with previous reports [@kitaoka2000; @kitaoka2007], suggesting an acute angle contraction effect at very small - as well as at sufficiently large angles (below 10 degrees for the former and between 50 to 90 degrees for the latter) between the target horizontal line and the biasing horizontal bars when the illusion strength is weak.

@@

```{r message=FALSE, warning=FALSE, fig.cap="The upper plots show the distribution of demographic and dispositional variables. The middle plots show the relationship between illusion sensitivity scores, sex, and age (solid lines indicate significant relationships). Bottom plots show the correlation between the general factor of illusion sensitivity (Factor \\textit{i}), and personality traits.", out.width="100%"}
knitr::include_graphics("figures/Figure3.png")
```

@@

We report significant links between inter-individual indices of illusion sensitivity and variables related to experimental context, demographic characteristics and personality. Firstly, screen size was found to have a significant negative relationship with the sensitivity to the two contrast-based illusions, namely the White and Contrast illusions. One possible explanation can be found in the mechanism by which visual systems filter through more low spatial frequencies when the size of the target object is small [@dixon2014]. As this filtering process excludes illumination information from visual processing, smaller screen sizes could yield artifactual changes in brightness perception, which in turn could attenuate the illusory effect of luminance-related illusions.
